---
title: "Analysis"
format: html
editor: visual
author: "Hiruni Daas Bandara"
---

# 0.1 Dependencies

```{r}
setwd("/path/to/project/folder")

library(reshape2)
library(ggplot2)
library(dplyr)

colanno <- read.csv('metadata/colanno27.csv', row.names=1)
colPal <- c('#12436D', '#28A197', '#801650', '#F46A25', '#3D3D3D', '#A285D1', 'rosybrown4',  'slategray4')
colPal_pie <- c('#12436D', '#28A197', 'slategray4', '#F46A25',  '#A285D1', 'rosybrown4', '#801650', '#3D3D3D')
```

# 0.2. Functions

```{r}
# aggregated matrix across the rows/cols; from an array/ list of matrices
listToMeanF <- function(m2, rows, cols){
  maxmat <- do.call(cbind, lapply(unique(cols), function(i){
    do.call(rbind, lapply(unique(rows), function(j){
      mean(m2[rows==j, cols==i,], na.rm=TRUE)
    }))
  }))
  rownames(maxmat) <- unique(rows)
  colnames(maxmat) <- unique(cols)
  maxmat
}

# scaling for the maximum on a column
scaleToMaxF <- function(mat){
  mat/ rep(matrixStats::colMaxs(mat), each=nrow(mat))
}

# heatmap for column-scaled matrix
meanHMplotF <- function(mat, title=NULL, row_groups=NULL, col_groups=NULL){
  df <- melt(mat)
  p <- ggplot(df, aes(x=Var2, y=Var1, fill = value)) +
    geom_tile() +
    geom_text(aes(label=round(value, 2)), color='red', size=3) +
    ylab('True Cell Type') +
    xlab('True Cell Type of Target Read') +
    scale_fill_viridis_c(option = 'B', direction = -1) +
    ggtitle(title) +
    guides(fill = guide_legend(title = "Probability (scaled to max)")) +
    theme(axis.text.x = element_text(size = 15, angle = 270, vjust = 1, hjust=1),
        axis.text.y = element_text(size = 15),
        axis.title = element_text(size = 18),
        legend.title = element_text(size=15),
        legend.text = element_text(size=15),
        plot.title = element_text(size = 20),
        rect = element_rect(fill = 'transparent'))


  if (!is.null(row_groups)) {
    # row_groups = named vector or list giving group sizes in order
    # e.g. c(GroupA = 3, GroupB = 4, GroupC = 2)
      breaks <- cumsum(row_groups[-length(row_groups)]) + 0.5
      p <- p + geom_hline(yintercept = breaks, color = "white", linewidth = 0.7)
    }
    
  # Add vertical separators between column groups
    if (!is.null(col_groups)) {
      breaks <- cumsum(col_groups[-length(col_groups)]) + 0.5
      p <- p + geom_vline(xintercept = breaks, color = "white", linewidth = 0.7)
    }
  p
}


```

# 1. Averaged performance of model

## 1.1 Comparison of target and reference reads (Step 3)

```{r}
# output
prodArr <- readRDS('output/2_prodArr_p49.rds')
prodArr1 <- readRDS('output/2_prodArr_p51.rds')
prodArr2 <- readRDS('output/2_prodArr_p54.rds')

# re-labeling
ctLabels <- c(paste(c(rep("T-CD4", 2), rep('T-CD8', 4)), c(5,6,1,2,3,4), sep='_'), colnames(prodArr)[7:30])
ctLabels_new <- paste0(stringr::str_split_i(ctLabels, '_', 1), '_', seq(1:30))
targetCT <- stringr::str_split_i(ctLabels, '_', 1)


mat_prod_list <- lapply(list(prodArr, prodArr1, prodArr2), function (arr){
    listToMeanF(arr, rownames(arr), colnames(arr))
})
matProdMean <- apply(simplify2array(mat_prod_list), c(1,2), mean)
scaledMean <- scaleToMaxF(matProdMean)[,ctLabels]
rownames(scaledMean) <- stringr::str_split_i(rownames(scaledMean), '_', 1)
colnames(scaledMean) <- ctLabels_new

# plot (Fig1.B)
meanHMplotF(scaledMean, col_groups = table(targetCT)[unique(targetCT)])
```

## 1.2 Reference prior (Step 2)

```{r}
refPriorWeights <- readRDS('output/2.1_localWeightsArr_perSubCT_rowColNorm_10kb.rds')
sctype_ref_mod <- c('T-CD4', 'T-CD8', 'T-CD4', unique(colanno[, 'stype'])[4:9])
sctype_ref <- rep(sctype_ref_mod, each=3)
mat_ref <- listToMeanF(refPriorWeights, sctype_ref, colnames(refPriorWeights))

# plot (Fig1.C)
meanHMplotF(scaleToMaxF(mat_ref))
```

## 1.3 Weighted output (Step 1)

```{r}
eq2 <- readRDS('output/3_weightedProdEQ2/weightedProd_6CT_p49_10kb.rds')
eq2.1 <- readRDS('output/3_weightedProdEQ2/weightedProd_6CT_p51_10kb.rds')
eq2.2 <- readRDS('output/3_weightedProdEQ2/weightedProd_6CT_p54_10kb.rds')

mat_eq2_list <- lapply(list(eq2, eq2.1, eq2.2), function(arr){
  listToMeanF(arr,  rownames(eq2), colnames(eq2))
})

meanEq2 <- apply(simplify2array(mat_eq2_list), c(1,2), mean)
meanEq2 <- meanEq2[,ctLabels]
colnames(meanEq2) <- ctLabels_new

# plot (Fig1.D)
meanHMplotF(scaleToMaxF(meanEq2), col_groups = table(targetCT)[unique(targetCT)])
```

# 2. Performance analysis

## 2.1 CT proportions

```{r}
eq2 <- readRDS('output/3_weightedProdEQ2/weightedProd_6CT_p49_10kb.rds')
cts <- stringr::str_split_i(colnames(eq2), '_', 1)

ctProb <- data.frame(table(cts)/length(cts) )
ctProb['Freq'] <- c(0.03, 0.44, 0.13, 0.20, 0.07, 0.13)
ctProb['Classification'] <- 'multi'
binary <- data.frame(
  'cts' = c('Lymphoid', 'Myeloid'),
  'Freq' = c(0.44, 0.56),
  'Classification' = 'binary'
)
meta <- rbind(ctProb, binary)

# plot (Fig2.A)
ggplot(meta, aes(x=Classification, y=Freq, fill=cts)) +
  geom_col() +
  coord_polar(theta = 'y') +
  scale_fill_manual(values = colPal_pie) +
  geom_text(label='0.38', x=1, y=0.76, size=4) +
  geom_text(label='0.62', x=0.2, y=0.8, size=4) +
  geom_text(data=subset(meta, Classification=='multi'), aes(label=as.character(Freq), x=Classification,  y=Freq), position = position_stack(vjust = 0.5), size=4)+
  ggtitle('Simulated Read Proportions by CT') +
  theme(
    axis.title=element_blank(),
    axis.ticks=element_blank(),
    axis.text=element_blank(),
    legend.title = element_text(size=15),
    legend.text = element_text(size=15),
    plot.title = element_text(size = 20),
    rect = element_rect(fill = 'transparent'),
    plot.background = element_rect(fill = "white", color = NA),
    panel.background = element_rect(fill = "white", color = NA),
    legend.background = element_rect(fill = "white", color = NA),
    legend.key = element_rect(fill = "white", color = NA)
  )
    
```

## 2.2 Threshold variation in probability margin

```{r}
data <- readRDS('output/4_weightedProdEQ2_6CT_confMat_10kb.rds')

cols <- colnames(data)[3:ncol(data)]
df <- as.data.frame(data) %>%
  mutate(across(all_of(cols), as.numeric)) 

# global mean
dfs <- split(df[,c(2:ncol(df))], df$pId)
cls <- unique(df$Class)

threshList <- lapply(dfs, function(i) {
  i <- dfs[[1]]
  if (length(cls)==2){ # binary
    sub <- i[i$Class == 'Lymphoid',]
    
    sapply(unique(i$Threshold), function(t) {
      mat <- sub[sub$Threshold==t, c(3:6)]
      mat <- mat/ rowSums(mat)
      (mat[,'TP'] + mat[,'TN']) / rowSums(mat)
    }, simplify = T)
    
  } else {# multiclass
    
    sapply(unique(i$Threshold), function(t) {
      mat <- i[i$Threshold==t, c(3:6)]
      mat <- mat/ rowSums(mat)
      sum(mat[,'TP']) / sum(mat[,c('TP', 'FN')])
    }, simplify = T)
  }
})

meanThresh <- rowMeans(matrix(unlist(threshList), ncol=3))
drop <- lapply(dfs, function(i){
  i[i$Class==cls[1], 'Dropout']
})
dropMeans <- rowMeans(matrix(unlist(drop), ncol=3))


dfplot <- data.frame(
  "Accuracy" = meanThresh,
  'Threshold' = unique(df$Threshold),
  'Dropout' = dropMeans
)

df_plot_sub <- filter(df_plot, Class=='All')

# plot (Fig2.B)
ggplot(data = df_plot_sub, aes(x=Threshold, y=Percent_Accuracy, fill = Category))+
  geom_bar(data = subset(df_plot_sub, Category!='Dropout'), position = 'fill', stat='identity') +
  geom_line(data = subset(df_plot_sub, Category=='Dropout'), linetype='dashed', colour='brown4', linewidth=0.8, alpha=0.8) +
  annotate('text', label='Dropout', x=1.2, y=0.03, size=5, color='brown4') +
  scale_x_continuous(breaks = unique(df_plot_sub$Threshold)) +
  ylab('Proportion of Reads') +
  coord_cartesian(clip = "off") +
  #scale_fill_viridis_d(option='cividis') +
  scale_fill_manual(values= c('Correct' = '#1a305e', 'Wrong'= 'gray40'))+
  #scale_fill_brewer(palette = 'PuBuGn') + #
  geom_hline(yintercept = 1/6, alpha=0.8, linewidth=0.8) +
  annotate('text', x=1.3, y=0.2, label='Random guessing', size=5) +
  ggtitle('Classification Accuracy against Read Dropout') +
  theme_minimal() +
  theme(axis.text.x = element_text(size = 15, angle = 270, vjust = 1, hjust=1),
        axis.text.y = element_text(size = 15),
        axis.title = element_text(size = 18),
        legend.title = element_text(size=15),
        legend.text = element_text(size=15),
        plot.title = element_text(size = 20),
        rect = element_rect(fill = 'transparent'))

```

## 2.3 Read Classification Accuracy

```{r}
data <- readRDS('output/4_weightedProdEQ2_6CT_confMat_10kb.rds')

cols <- colnames(data)[3:ncol(data)]
df <- as.data.frame(data) %>%
  mutate(across(all_of(cols), as.numeric)) 

# global mean
dfs <- split(df[,c(2:ncol(df))], df$pId)
cls <- unique(df$Class)

metrics <- sapply(dfs, function(i){
  mat <- i[i$Threshold == 1,]
  if (length(cls)==2){ # binary
    sub <- mat[mat$Class == 'Lymphoid', c(3:6)]
    sub <- sub/ rowSums(sub)
    acc <- (sub[,'TP'] + sub[,'TN']) / rowSums(sub)
    
  } else{# multiclass
    mat <- mat[,c(3:6)]
    mat <- mat/ rowSums(mat)
    acc <- sum(mat[,'TP']) / sum(mat[,c('TP', 'FN')])
  }
  precision <- mat[,'TP'] / (mat[,'TP'] + mat[,'FP'])
  recall <- mat[,'TP'] / (mat[,'TP'] + mat[,'FN'])
  f1 <- (2*precision*recall) / (precision+recall)

  c(acc, precision, recall, f1)
})
accMean <- rowMeans(metrics)
metric <- c('Accuracy', rep(c('Precision', 'Recall', 'F1'), each=length(cls)))

dfplot <- data.frame(
  'Metric'= metric,
  'Class' = c('Overall', rep(cls, 3)),
  'Value' = accMean*100
  )
dfplot$Class <- factor(dfplot$Class, levels = c(cls, 'Overall')) 

# plot (Fig2.C/D)
ggplot(dfplot, aes(x=Class, y=Value, fill=Metric)) +
  geom_bar(stat='identity',
           position = position_dodge2(width = 0.9, preserve = "single"))+
  geom_hline(yintercept = 1/6*100, alpha=0.8, linewidth=0.8, linetype='dashed', color='black') +
  annotate('text', x=8, y=19, label='Random guessing', size=5) +
## For lymphoid vs myeloid:
  # geom_hline(yintercept = 1/2*100, alpha=0.8, linewidth=0.8, linetype='dashed', color='black') +
  # annotate('text', x=3.9, y=53, label='Random guessing', size=5) +
  scale_fill_viridis_d(option='cividis') +
  ylab('Percentage') +
  coord_cartesian(clip = "off") +
  ggtitle('Classification Performance of 10kb Reads (6CT)') +
  theme_minimal() +
  theme(axis.text.x = element_text(size = 15, angle = 270, vjust = 1, hjust=1),
        axis.text.y = element_text(size = 15),
        axis.title = element_text(size = 18),
        legend.title = element_text(size=15),
        legend.text = element_text(size=15),
        plot.title = element_text(size = 20),
        rect = element_rect(fill = 'transparent'))

```

# 3. Classification Accuracy across the Genome

## 3.1 Performance at 6CT and 2CT classification

```{r}
data <- readRDS('output/4_weightedProdEQ2_genome_6CT_stats.rds')
threshold <- 1

cols <- colnames(data)[3:ncol(data)]
df <- as.data.frame(data) %>%
  mutate(across(all_of(cols), as.numeric))

# global mean
dfs <- split(df[,c(2:ncol(df))], df$pId)

threshArr <-lapply(dfs, function(i) {
  sub <- i[i$Threshold==threshold,]
  sub[,5:ncol(sub)] <- sub[,5:ncol(sub)] / sub[,4]
})
meanMats <- Reduce('+', threshArr) / length(threshArr)

meta <- dfs[[1]][dfs[[1]]$Threshold==threshold, c(1:4)]
percentData <- bind_cols(meta, meanMats) 
dfplot <- percentData %>%
  mutate(Threshold=as.character(Threshold)) %>%
  tidyr::pivot_longer(c(5:ncol(percentData)), names_to = 'Min_accuracy', values_to='Window_prop')

dfplot$Min_accuracy <- factor(dfplot$Min_accuracy, levels=c('Null', setdiff(dfplot$Min_accuracy, 'Null')))
dfplot$Class <- factor(dfplot$Class, levels = unique(dfplot$Class))

# plot (Fig3.A/B)
ggplot(data = dfplot, aes(x=Min_accuracy, 
                                y=Window_prop, colour = Class, group=Class))+
  geom_line(linewidth=1.1, alpha=0.6) +
  xlab('Recall of a given 10kb Region (<) ') +
  ylab('Proportion of Autosomal Genome') +
  ggtitle('Cell Type Recall over the Genome') +
  scale_colour_manual(values=c('goldenrod3', colPal[c(3,5)])) +
  theme_minimal() +
  theme(axis.text.x = element_text(size = 15, angle = 270, vjust = 1, hjust=1),
        axis.text.y = element_text(size = 15),
        axis.title = element_text(size = 18),
        legend.title = element_text(size=15),
        legend.text = element_text(size=15),
        plot.title = element_text(size = 20),
        rect = element_rect(fill = 'transparent'))

```

## 3.2 Read size comparison

```{r}
acc_genome_10 <- readRDS('output/4_weightedProdEQ2_genome_LymMyl_stats.rds')
acc_genome_15 <- readRDS('output/4_weightedProdEQ2_genome_LymMyl_stats_15kb.rds')
acc_genome_20 <- readRDS('output/4_weightedProdEQ2_genome_LymMyl_stats_20kb.rds')

genomeList <- list(acc_genome_10, acc_genome_15, acc_genome_20)
strLen <- c('10', '15', '20')
meanStr <- list()
for (i in seq(length(genomeList))) {
  data <- genomeList[[i]]

  cols <- colnames(data)[3:ncol(data)]
  df <- as.data.frame(data) %>%
    mutate(across(all_of(cols), as.numeric)) %>%
    filter(Threshold==1) 
  
  # global mean
  dfs <- split(df[,c(2:ncol(df))], df$pId)
  
  matlist <- lapply(dfs, function(sub){
    sub[,5:ncol(sub)] <- sub[,5:ncol(sub)] / sub[,4]
  })

  meanMats <- Reduce('+', matlist) / length(matlist)
  
  sdMat <- sapply(matlist, function(mat){mat[,'0.9']}, simplify = T)
  sdMat <- apply(sdMat, 1, sd)
  meanMats <- dfs[[1]] %>%
    select(Class, Total_nslices_left) %>%
    mutate('strLen'= strLen[i]) %>%
    mutate('sd' = sdMat) %>%
    bind_cols(meanMats) 
  meanStr[[i]] <- meanMats
}

meanStrAll <- do.call(rbind, meanStr)

dfplot <- meanStrAll %>%
  select(c(1,3,4,15)) %>%
  rename('Read Length' = strLen)

# plot (Fig3.C)
ggplot(dfplot, aes(x=Class, y=`0.9`, group=`Read Length`, fill=`Read Length`)) +
  geom_bar(position = 'dodge', stat='identity') +
  scale_fill_brewer(palette = 'Paired') +
  ggtitle('Classification Recall (>0.9) & Read Length') +
  ylab('Proportion of Regions (Recall > 0.9) ') +
  theme_minimal() +
  theme(axis.text.x = element_text(size = 15, angle = 270, vjust = 1, hjust=1),
        axis.text.y = element_text(size = 15),
        axis.title = element_text(size = 18),
        legend.title = element_text(size=15),
        legend.text = element_text(size=15),
        plot.title = element_text(size = 20),
        rect = element_rect(fill = 'transparent'))

```
